---
title: "Statistical Testing in Health Sciences"
subtitle: "A Framework for Avoiding Errors in Analysis"
author: "Dr Simon Thornley"
date: "January 6, 2026"
format:
  pptx:
    slide-level: 2
prefer-html: true
knitr:
  opts_chunk:
    dev: "png"
editor: visual
---

## Learning Objectives

By the end of this presentation, you will:

- Understand why statistical testing matters in health research
- Learn a systematic framework for data analysis
- Recognize common errors and how to avoid them
- Apply basic statistical tests in R
- Interpret results appropriately

## Why Statistical Testing Matters

**Real-world consequences:**

- Wrong conclusions about treatment effectiveness
- Wasted research funding and time
- Patient safety implications
- Reproducibility crisis in science

**Example:** A study claims a new drug works, but errors in analysis led to false positives. Patients receive ineffective treatment.

## The Analysis Framework

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 5
library(DiagrammeR)
grViz("
digraph analysis_flow {
  node [shape=box, style=filled, fillcolor=lightblue]
  Import -> Clean -> Explore -> Test -> Interpret
  
  Clean -> Import [label='Issues?', style=dashed]
  Explore -> Clean [label='Problems?', style=dashed]
  Test -> Explore [label='Assumptions?', style=dashed]
}
", width = 800, height = 300)
```

**Five key steps - each with potential pitfalls!**

# Step 1: Data Import

## Import: Common Errors

**What can go wrong?**

1. Wrong delimiter (comma vs. tab vs. semicolon)
2. Column names not read correctly
3. Missing values coded inconsistently (NA, -999, blank)
4. Date formats not recognized
5. Special characters cause issues

## Import: Good Practice

```{r}
#| echo: true
#| eval: false
# Load necessary packages
library(readr)
library(dplyr)

# Import with explicit specifications
data <- read_csv(
  "patient_data.csv",
  col_types = cols(
    patient_id = col_character(),
    age = col_double(),
    treatment = col_factor(),
    outcome = col_double()
  ),
  na = c("", "NA", "N/A", "-999")
)

# ALWAYS check your import!
glimpse(data)
head(data)
summary(data)
```

## Import: Red Flags to Check

```{r}
#| echo: true
#| eval: false
# Check for problems
problems(data)  # Any parsing issues?

# Verify dimensions
dim(data)  # Expected rows and columns?

# Check column types
str(data)  # Are numeric columns actually numeric?

# Look for unexpected values
unique(data$treatment)  # Any typos?
```

**Stop and fix issues NOW - don't continue with bad data!**

# Step 2: Data Cleaning

## Cleaning: Common Errors

**Critical mistakes:**

1. Not checking for duplicates
2. Ignoring missing data patterns
3. Removing outliers without justification
4. Not documenting changes
5. Overwriting original data

## Cleaning: Example Dataset

```{r}
#| echo: true
# Create example health study data
set.seed(123)
n <- 200

patient_data <- data.frame(
  id = 1:n,
  age = round(rnorm(n, 45, 15)),
  sex = sample(c("M", "F"), n, replace = TRUE),
  treatment = sample(c("Control", "Drug A", "Drug B"), n, replace = TRUE),
  baseline_bp = round(rnorm(n, 140, 20)),
  followup_bp = round(rnorm(n, 130, 20)),
  adverse_event = sample(c("Yes", "No"), n, replace = TRUE, prob = c(0.15, 0.85))
)

# Introduce some realistic problems
patient_data$age[c(5, 23)] <- NA
patient_data$followup_bp[c(10, 45, 67)] <- NA
patient_data$id[150] <- 149  # Duplicate
```

## Cleaning: Systematic Approach

```{r}
#| echo: true
library(dplyr)

# 1. Check for duplicates
duplicates <- patient_data %>% 
  group_by(id) %>% 
  filter(n() > 1)

if(nrow(duplicates) > 0) {
  warning("Duplicates found!")
  print(duplicates)
}

# 2. Check missing data
colSums(is.na(patient_data))

# 3. Check for impossible values
summary(patient_data$age)  # Negative ages?
summary(patient_data$baseline_bp)  # BP = 0 or 300?
```

## Cleaning: Handling Missing Data

```{r}
#| echo: true
# Visualize missing patterns
library(naniar)
vis_miss(patient_data)

# Decide on strategy:
# - Complete case analysis (if <5% missing at random)
# - Imputation (if defensible)
# - Report as limitation

# Document your decision!
clean_data <- patient_data %>%
  filter(!duplicated(id)) %>%  # Remove duplicates
  filter(!is.na(age) & !is.na(followup_bp))  # Complete cases

cat("Removed", nrow(patient_data) - nrow(clean_data), "rows")
```

# Step 3: Exploratory Analysis

## Exploration: Why It Matters

**Before statistical testing, you MUST:**

- Understand your data distribution
- Check assumptions
- Identify outliers and patterns
- Choose appropriate tests

**Skipping exploration = guaranteed errors**

## Exploration: Contingency Tables

```{r}
#| echo: true
# Cross-tabulation
table(clean_data$treatment, clean_data$adverse_event)

# Add proportions
prop.table(table(clean_data$treatment, clean_data$adverse_event), 
           margin = 1) %>% 
  round(3)

# More informative summary
library(janitor)
clean_data %>%
  tabyl(treatment, adverse_event) %>%
  adorn_percentages("row") %>%
  adorn_pct_formatting(digits = 1) %>%
  adorn_ns()
```

## Exploration: Visualize Before Testing!

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 5
library(ggplot2)

ggplot(clean_data, aes(x = treatment, y = baseline_bp)) +
  geom_boxplot(fill = "lightblue") +
  geom_jitter(width = 0.2, alpha = 0.3) +
  labs(title = "Baseline Blood Pressure by Treatment Group",
       y = "Systolic BP (mmHg)") +
  theme_minimal()
```

## Exploration: Check Distributions

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 4
library(patchwork)

p1 <- ggplot(clean_data, aes(x = age)) +
  geom_histogram(bins = 30, fill = "steelblue") +
  labs(title = "Age Distribution") +
  theme_minimal()

p2 <- ggplot(clean_data, aes(sample = age)) +
  stat_qq() + stat_qq_line() +
  labs(title = "Q-Q Plot: Normality Check") +
  theme_minimal()

p1 + p2
```

# Step 4: Statistical Testing

## Testing: Choose the Right Test

**Common mistake: Using the wrong test!**

| Data Type | Groups | Test |
|-----------|--------|------|
| Continuous | 2 groups | t-test (or Wilcoxon) |
| Continuous | 3+ groups | ANOVA (or Kruskal-Wallis) |
| Categorical | 2+ groups | Chi-square |
| Continuous outcome + predictors | Multiple | Regression |

**Check assumptions first!**

## Testing: Chi-Square Example

```{r}
#| echo: true
# Question: Is adverse event rate different by treatment?

# Create contingency table
cont_table <- table(clean_data$treatment, clean_data$adverse_event)
cont_table

# Chi-square test
chi_result <- chisq.test(cont_table)
chi_result

# Check expected frequencies (should be >5)
chi_result$expected
```

## Testing: Interpreting Chi-Square

```{r}
#| echo: true
# Extract key information
cat("Chi-square statistic:", round(chi_result$statistic, 2), "\n")
cat("Degrees of freedom:", chi_result$parameter, "\n")
cat("P-value:", round(chi_result$p.value, 4), "\n")

if(chi_result$p.value < 0.05) {
  cat("\nConclusion: Significant difference in adverse event rates")
} else {
  cat("\nConclusion: No significant difference in adverse event rates")
}
```

**Important:** P-value is not effect size! Report both.

## Testing: T-Test Example

```{r}
#| echo: true
# Question: Does Drug A reduce BP more than Control?

# Calculate BP change
clean_data <- clean_data %>%
  mutate(bp_change = baseline_bp - followup_bp)

# Filter to two groups
compare_data <- clean_data %>%
  filter(treatment %in% c("Control", "Drug A"))

# Check normality assumption
shapiro.test(compare_data$bp_change[compare_data$treatment == "Control"])
shapiro.test(compare_data$bp_change[compare_data$treatment == "Drug A"])
```

## Testing: T-Test Results

```{r}
#| echo: true
# Perform t-test
t_result <- t.test(bp_change ~ treatment, data = compare_data)
t_result

# Calculate effect size (Cohen's d)
library(effsize)
cohen.d(bp_change ~ treatment, data = compare_data)
```

**Always report effect size with p-value!**

## Testing: Linear Regression

```{r}
#| echo: true
# Question: What predicts BP reduction?

model <- lm(bp_change ~ age + sex + treatment + baseline_bp, 
            data = clean_data)

summary(model)
```

## Testing: Check Regression Assumptions

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 6
par(mfrow = c(2, 2))
plot(model)
```

**Red flags:** Patterns in residuals, outliers with high leverage

## Testing: Regression Interpretation

```{r}
#| echo: true
# Extract coefficients with confidence intervals
confint(model)

# More readable output
library(broom)
tidy(model, conf.int = TRUE) %>%
  select(term, estimate, conf.low, conf.high, p.value) %>%
  mutate(across(where(is.numeric), round, 3))
```

**Interpretation:** After controlling for other factors, each 1 mmHg higher baseline BP is associated with X mmHg greater reduction.

# Step 5: Visualization

## Plots: Principles

**Good statistical plots:**

1. Show the data (not just summaries)
2. Include error bars/confidence intervals
3. Label everything clearly
4. Choose appropriate scale
5. Don't mislead with truncated axes

## Plots: Before-After Comparison

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 5
library(tidyr)

plot_data <- clean_data %>%
  select(id, treatment, baseline_bp, followup_bp) %>%
  pivot_longer(cols = c(baseline_bp, followup_bp),
               names_to = "timepoint",
               values_to = "bp")

ggplot(plot_data, aes(x = timepoint, y = bp, group = id)) +
  geom_line(alpha = 0.2) +
  geom_point(alpha = 0.3) +
  stat_summary(aes(group = 1), fun = mean, geom = "line", 
               color = "red", size = 1.5) +
  facet_wrap(~treatment) +
  labs(title = "Blood Pressure Change by Treatment",
       y = "Systolic BP (mmHg)") +
  theme_minimal()
```

## Plots: Effect Sizes with CI

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 5
# Calculate mean change and CI by group
summary_data <- clean_data %>%
  group_by(treatment) %>%
  summarise(
    mean_change = mean(bp_change, na.rm = TRUE),
    se = sd(bp_change, na.rm = TRUE) / sqrt(n()),
    ci_lower = mean_change - 1.96 * se,
    ci_upper = mean_change + 1.96 * se
  )

ggplot(summary_data, aes(x = treatment, y = mean_change)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
  labs(title = "Mean BP Reduction by Treatment (with 95% CI)",
       y = "BP Change (mmHg)") +
  theme_minimal()
```

# Common Errors to Avoid

## Top 10 Statistical Mistakes

1. **P-hacking:** Testing multiple hypotheses, only reporting significant ones
2. **Cherry-picking:** Removing "inconvenient" data without justification
3. **Wrong test:** Not checking assumptions
4. **Ignoring multiple comparisons:** Need correction for many tests
5. **Confusing significance with importance:** p<0.05 ≠ clinical relevance

## Top 10 Statistical Mistakes (cont.)

6. **Assuming causation from correlation:** Association ≠ causation
7. **Inadequate sample size:** Underpowered studies miss true effects
8. **Not reporting effect sizes:** P-value alone is insufficient
9. **Ignoring missing data:** Can bias results
10. **Not documenting analysis:** Can't reproduce = not science

## The Reproducibility Checklist

**Before you publish/present:**

- [ ] Raw data preserved
- [ ] Analysis code documented
- [ ] Random seeds set (for reproducibility)
- [ ] All exclusions justified
- [ ] Assumptions checked
- [ ] Effect sizes reported
- [ ] Confidence intervals included
- [ ] Multiple testing addressed
- [ ] Limitations acknowledged

# Practical Example: Full Analysis

## Complete Workflow Example

```{r}
#| echo: true
#| eval: false
# 1. IMPORT
library(readr)
data_raw <- read_csv("study_data.csv")

# 2. CLEAN
data_clean <- data_raw %>%
  filter(!duplicated(id)) %>%
  filter(complete.cases(.)) %>%
  mutate(treatment = factor(treatment))

# 3. EXPLORE
summary(data_clean)
ggplot(data_clean, aes(x = outcome)) + geom_histogram()

# 4. TEST
model <- lm(outcome ~ treatment + age + sex, data = data_clean)
summary(model)

# 5. VISUALIZE
ggplot(data_clean, aes(x = treatment, y = outcome)) +
  geom_boxplot() + theme_minimal()
```

## Your Analysis Template

```{r}
#| echo: true
#| eval: false
# Copy this structure for your assignments!

# Load packages
library(tidyverse)
library(broom)

# 1. Import and inspect
data <- read_csv("yourdata.csv")
glimpse(data)

# 2. Clean and document
data_clean <- data %>%
  filter(___) %>%  # Document why
  mutate(___)      # Create needed variables

# 3. Explore first!
summary(data_clean)
# Create plots to understand distributions

# 4. Test appropriately
# Check assumptions
# Run test
# Check effect size

# 5. Visualize results
# Make it clear and honest
```

# Key Takeaways

## Remember These Principles

1. **Check your data at every step** - errors compound!
2. **Visualize before testing** - understand before concluding
3. **Choose tests based on assumptions** - not wishful thinking
4. **Report effect sizes and confidence intervals** - not just p-values
5. **Document everything** - reproducibility matters

## Statistical Testing ≠ Magic

**Statistics can't:**

- Fix bad study design
- Make insufficient data sufficient  
- Prove causation from correlation
- Replace domain expertise

**Statistics can:**

- Quantify uncertainty
- Test hypotheses rigorously
- Identify patterns in data
- Support evidence-based decisions

## Resources for Learning More

**Books:**
- "The Art of Statistics" - David Spiegelhalter
- "Statistical Rethinking" - Richard McElreath

**Online:**
- R for Data Science (r4ds.had.co.nz)
- Cross Validated (stats.stackexchange.com)

**Practice:**
- Work through examples
- Check your assumptions
- Compare with classmates
- Ask questions!

## Questions?

**Key message:** Good statistical analysis requires:

- **Skepticism** - question every step
- **Transparency** - document everything
- **Humility** - acknowledge limitations
- **Rigor** - follow best practices

**Your health research conclusions affect real people - do it right!**
